Investigating the impact of interventionist causal approach on the study of verbal aggression and discrimination online

### 2021-09-08 paperDraft/paperDraft.Rmd
some paragraphs added about causality from old writings - it's still chaotic though, need to improve

# Section Abstract
This article describes an experimental intervention study conducted in a naturalistic, digital setting (Q&A forum on Reddit), utilizing a collective intelligence approach to content moderation and reduction of the level of verbal aggression among a selected group of Reddit users who regularly attack other community members. Collective Intelligence in this sense means exploring the collaboration between human and machine intelligence to develop solutions to social challenges. Artificial Intelligence was used to detect verbal aggression (personal attacks) and notify human volunteers about attacks. Volunteers after receiving notifications employed interventions based on norm or empathy promotion. We find that only those who were sanctioned with norms-inducing interventions had their personal attacks' level significantly decreased.

@@

# Introduction
Although much effort has been made in order to tackle the problem of verbal aggression and harassment online, looking at various reports and surveys [@sorrentino2019epidemiology; @vogels_state_2021; @Zachary_hate], it remains a common hindrance for people engaging with social media in their everyday lives. The problem got exacerbated in the midst of the COVID19 pandemic, during which the majority of our social life moved to cyberspace. During this shift, there was an increase in cyberbullying attitudes and perpetration [@barlett2021comparing], 90% increase in public reports of illegal online content [@grant_2021], including 114% increase in non-consensual sharing of intimate images, 30% increase in cyberbullying, as well as 40% of increase in adults reporting online harassment. According to a report conducted by company L1ght [@noauthor_l1ght_2020], hate speech directed towards China and the Chinese went up by 900% on Twitter. Gaming platforms were in the spotlight as well, with a 40% increase in toxicity on Discord \footnote{In both reports, the increase is reported as a relative change between the year 2019 and 2020, with no absolute indicators.}.

>>>>

# Section Abstract 
This article \todo{hm not sure whether abstract is neccessary in BA thesis but I will leave it here for now} describes an experimental intervention study conducted in a naturalistic, digital setting (Q&A forum on Reddit), utilizing a collective intelligence approach to content moderation and reduction of the level of verbal aggression among a selected group of Reddit users who regularly attack other community members. Collective Intelligence in this sense means exploring the collaboration between human and machine intelligence to develop solutions to social challenges. Artificial Intelligence was used to detect verbal aggression (personal attacks) and notify human volunteers about attacks. Volunteers after receiving notifications employed interventions based on norm or empathy promotion. We find that only those who were sanctioned with norms-inducing interventions had their personal attacks' level significantly decreased.

@@

# Introduction

We live in a world in which there is an abundance of data. Paraphrasing the phrase coined by Clive Humby, "data is the new oil" --- data has displaced oil and become the world’s most valuable resource. Albeit there are some fundamental differences between the two assets. Oil belongs to a very limited and finite resource, while data is growing at an exponential rate (cite: https://cloudtweaks.com/2015/03/how-much-data-is-produced-every-day/). The value of oil does not depend on its amount or the refinery processes - it's rather constant at a particular dynamic of market's supply and demand. Data and its value, on the contrary, is inherently tied to its amount and the "refinery process" - various tools and techniques for its discovery and analysis. In some cases, certain techniques can be used only if one has a sufficient (meaning enormous) data at one's disposal --- e.g. statistical methods like machine learning and deep learning \todo{@article{lecun2015deep,
  title={Deep learning},
  author={LeCun, Yann and Bengio, Yoshua and Hinton, Geoffrey},
  journal={nature},
  volume={521},
  number={7553},
  pages={436--444},
  year={2015},
  publisher={Nature Publishing Group}
}}. 
In the last couple of years, more and more researchers started to research and developed tools for causal discovery and inference. Unfortunately, although thinking in causal terms seems to be our inherent human capacity and one of our basic cognitive abilities, that is not necessarily the case with the majority of toolkits and statistical models developed to draw insight and correlations from data. 
In classical statistics, which is stripped of causality-related terms and reasoning, one cannot speak about causal effects, unless one deals with a randomized controlled experiment. This comprises one of the limits of the classical paradigm in statistics (among others such as arbitrary thresholds of statistical significance, p-hacking, disregard of prior information, and others which we will not discuss here).
But often scientists conduct the so-called cross-sectional research, in which no variable is manipulated. In the language and toolkit of classical statistics in such cases we cannot conduct any causal inference. What we can measure using those methods is correlation, and, as frequently repeated, “correlation does not imply causation”. Correlation can be positive, negative, or there can be no correlation at all. A positive one implies that the increase of the value of one variable will influence the increase of another variables’ value. A negative correlation means that the change in values of two variables will diverge into separate directions --- one will increase and one will decrease alongside the first one. We can measure the strength of the correlation using various mathematical methods, where 1 or -1 means we are dealing with perfect correlation --- the value of one variable changes proportional to the value of another variable. But even though causation is necessarily connected with correlation, as already mentioned --- not every correlation implies causation. And as we will see, ignoring the causal story behind the data and relationships between the variables might lead to far-reaching consequences. Drawing false causal conclusions out of data can lead to false accusations of discrimination (...here I will add something more once I decide which examples to choose). This is especially problematic if we take into consideration the fact that statisticians and people in general naturally think in causal terms and spurious interpretations of studies occur frequently. On the other hand, a conservative approach of many classical statisticians led to a long-living debate on whether smoking causes cancer, lasting for over 20 years (to check). One of the reasons for such a fierce debate was the lack of a causal framework, which was still in its infancy at that time. Framework of tools with which we could validate or invalidate assumptions about the existence of the potential confounder - a vari- able causing both craving cigarettes and lung cancer (e.g. some genetic factor), creating a spurious correlation between cigarettes and lung cancer.
Modeling causality and moving beyond traditional statistics is a relatively young endeavor. One of the pioneers of the work on causality is Judea Pearl. As he describes, causality and traditional statistics are governed by opposing ways of thinking. While statistics might be described as “study of methods of the reduction of data” (quoted here was Fischer I think, need to check), causality requires the incorporation of an extra layer - the understanding or prior knowl- edge (assumptions) about the process that generate the data in the first place. Classical statistics might be described here as a “model-free approach,” just as Artificial Intelligence methods based on machine learning/deep learning, in which no prior knowledge can be taken into account.

In this paper, I will show how mathematical methods for causality modeling can be applied to discover how different variables affect each other, moving beyond traditional correlation. Data used for analysis comes from an experimental intervention study conducted in a naturalistic, digital setting (Q&A forum on Reddit), utilizing a collective intelligence approach to content moderation and reduction of the level of verbal aggression among a selected group of Reddit users who regularly attack other community members. Collective Intelligence in this sense means exploring the collaboration between human and machine intelligence to develop solutions to social challenges. Artificial Intelligence was used to detect verbal aggression (personal attacks) and notify human volunteers about attacks. Volunteers after receiving notifications employed interventions grounded in philosophy. There were two broad categories - one based on norms (deontology, virtue ethics among others) and second based on empathy (inspired by notions of David Hume and Adam Smith).

Although much effort has been made in order to tackle the problem of verbal aggression and harassment online, looking at various reports and surveys [@sorrentino2019epidemiology; @vogels_state_2021; @Zachary_hate], it remains a common hindrance for people engaging with social media in their everyday lives. The problem got exacerbated in the midst of the COVID19 pandemic, during which the majority of our social life moved to cyberspace. During this shift, there was an increase in cyberbullying attitudes and perpetration [@barlett2021comparing], 90% increase in public reports of illegal online content [@grant_2021], including 114% increase in non-consensual sharing of intimate images, 30% increase in cyberbullying, as well as 40% of increase in adults reporting online harassment. According to a report conducted by company L1ght [@noauthor_l1ght_2020], hate speech directed towards China and the Chinese went up by 900% on Twitter. Gaming platforms were in the spotlight as well, with a 40% increase in toxicity on Discord \footnote{In both reports, the increase is reported as a relative change between the year 2019 and 2020, with no absolute indicators.}.

### 2021-09-08 paperDraft/paperDraft.tex
some paragraphs added about causality from old writings - it's still chaotic though, need to improve

This article describes an experimental intervention study conducted in a
naturalistic, digital setting (Q\&A forum on Reddit), utilizing a
collective intelligence approach to content moderation and reduction of
the level of verbal aggression among a selected group of Reddit users
who regularly attack other community members. Collective Intelligence in
this sense means exploring the collaboration between human and machine
intelligence to develop solutions to social challenges. Artificial
Intelligence was used to detect verbal aggression (personal attacks) and
notify human volunteers about attacks. Volunteers after receiving
notifications employed interventions based on norm or empathy promotion.
We find that only those who were sanctioned with norms-inducing
interventions had their personal attacks' level significantly decreased.

@@

>>>>

This article
\todo{hm not sure whether abstract is neccessary in BA thesis but I will leave it here for now}
describes an experimental intervention study conducted in a
naturalistic, digital setting (Q\&A forum on Reddit), utilizing a
collective intelligence approach to content moderation and reduction of
the level of verbal aggression among a selected group of Reddit users
who regularly attack other community members. Collective Intelligence in
this sense means exploring the collaboration between human and machine
intelligence to develop solutions to social challenges. Artificial
Intelligence was used to detect verbal aggression (personal attacks) and
notify human volunteers about attacks. Volunteers after receiving
notifications employed interventions based on norm or empathy promotion.
We find that only those who were sanctioned with norms-inducing
interventions had their personal attacks' level significantly decreased.

@@

We live in a world in which there is an abundance of data. Paraphrasing
the phrase coined by Clive Humby, ``data is the new oil'' --- data has
displaced oil and become the world's most valuable resource. Albeit
there are some fundamental differences between the two assets. Oil
belongs to a very limited and finite resource, while data is growing at
an exponential rate (cite:
\url{https://cloudtweaks.com/2015/03/how-much-data-is-produced-every-day/}).
The value of oil does not depend on its amount or the refinery processes
- it's rather constant at a particular dynamic of market's supply and
demand. Data and its value, on the contrary, is inherently tied to its
amount and the ``refinery process'' - various tools and techniques for
its discovery and analysis. In some cases, certain techniques can be
used only if one has a sufficient (meaning enormous) data at one's
disposal --- e.g.~statistical methods like machine learning and deep
learning \todo{@article{lecun2015deep,
  title={Deep learning},
  author={LeCun, Yann and Bengio, Yoshua and Hinton, Geoffrey},
  journal={nature},
  volume={521},
  number={7553},
  pages={436--444},
  year={2015},
  publisher={Nature Publishing Group}
}}. In the last couple of years, more and more researchers started to
research and developed tools for causal discovery and inference.
Unfortunately, although thinking in causal terms seems to be our
inherent human capacity and one of our basic cognitive abilities, that
is not necessarily the case with the majority of toolkits and
statistical models developed to draw insight and correlations from data.
In classical statistics, which is stripped of causality-related terms
and reasoning, one cannot speak about causal effects, unless one deals
with a randomized controlled experiment. This comprises one of the
limits of the classical paradigm in statistics (among others such as
arbitrary thresholds of statistical significance, p-hacking, disregard
of prior information, and others which we will not discuss here). But
often scientists conduct the so-called cross-sectional research, in
which no variable is manipulated. In the language and toolkit of
classical statistics in such cases we cannot conduct any causal
inference. What we can measure using those methods is correlation, and,
as frequently repeated, ``correlation does not imply causation.''
Correlation can be positive, negative, or there can be no correlation at
all. A positive one implies that the increase of the value of one
variable will influence the increase of another variables' value. A
negative correlation means that the change in values of two variables
will diverge into separate directions --- one will increase and one will
decrease alongside the first one. We can measure the strength of the
correlation using various mathematical methods, where 1 or -1 means we
are dealing with perfect correlation --- the value of one variable
changes proportional to the value of another variable. But even though
causation is necessarily connected with correlation, as already
mentioned --- not every correlation implies causation. And as we will
see, ignoring the causal story behind the data and relationships between
the variables might lead to far-reaching consequences. Drawing false
causal conclusions out of data can lead to false accusations of
discrimination (\ldots here I will add something more once I decide
which examples to choose). This is especially problematic if we take
into consideration the fact that statisticians and people in general
naturally think in causal terms and spurious interpretations of studies
occur frequently. On the other hand, a conservative approach of many
classical statisticians led to a long-living debate on whether smoking
causes cancer, lasting for over 20 years (to check). One of the reasons
for such a fierce debate was the lack of a causal framework, which was
still in its infancy at that time. Framework of tools with which we
could validate or invalidate assumptions about the existence of the
potential confounder - a vari- able causing both craving cigarettes and
lung cancer (e.g.~some genetic factor), creating a spurious correlation
between cigarettes and lung cancer. Modeling causality and moving beyond
traditional statistics is a relatively young endeavor. One of the
pioneers of the work on causality is Judea Pearl. As he describes,
causality and traditional statistics are governed by opposing ways of
thinking. While statistics might be described as ``study of methods of
the reduction of data'' (quoted here was Fischer I think, need to
check), causality requires the incorporation of an extra layer - the
understanding or prior knowl- edge (assumptions) about the process that
generate the data in the first place. Classical statistics might be
described here as a ``model-free approach,'' just as Artificial
Intelligence methods based on machine learning/deep learning, in which
no prior knowledge can be taken into account.

In this paper, I will show how mathematical methods for causality
modeling can be applied to discover how different variables affect each
other, moving beyond traditional correlation. Data used for analysis
comes from an experimental intervention study conducted in a
naturalistic, digital setting (Q\&A forum on Reddit), utilizing a
collective intelligence approach to content moderation and reduction of
the level of verbal aggression among a selected group of Reddit users
who regularly attack other community members. Collective Intelligence in
this sense means exploring the collaboration between human and machine
intelligence to develop solutions to social challenges. Artificial
Intelligence was used to detect verbal aggression (personal attacks) and
notify human volunteers about attacks. Volunteers after receiving
notifications employed interventions grounded in philosophy. There were
two broad categories - one based on norms (deontology, virtue ethics
among others) and second based on empathy (inspired by notions of David
Hume and Adam Smith).
